---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# DBI

<!-- badges: start -->
[![Build Status](https://travis-ci.org/r-dbi/DBI.svg?branch=master)](https://travis-ci.org/r-dbi/DBI)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/r-dbi/DBI?branch=master&svg=true)](https://ci.appveyor.com/project/r-dbi/DBI)
[![Coverage Status](https://codecov.io/gh/r-dbi/DBI/branch/master/graph/badge.svg)](https://codecov.io/github/r-dbi/DBI?branch=master)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/DBI)](https://cran.r-project.org/package=DBI)
[![CII Best Practices](https://bestpractices.coreinfrastructure.org/projects/1882/badge)](https://bestpractices.coreinfrastructure.org/projects/1882)
<!-- badges: end -->

The DBI package helps connecting R to database management systems (DBMS).
DBI separates the connectivity to the DBMS into a "front-end" and a "back-end".
The package defines an interface that is implemented by *DBI backends* such as:

- [RPostgres](https://rpostgres.r-dbi.org),
- [RMariaDB](https://rmariadb.r-dbi.org),
- [RSQLite](https://rsqlite.r-dbi.org),
- [odbc](https://github.com/r-dbi/odbc),
- [bigrquery](https://github.com/r-dbi/bigrquery),

and many more.
R scripts and packages use DBI to access various databases through their DBI backends.

The interface defines a small set of classes and methods similar in spirit to Perl's [DBI](http://dbi.perl.org/), Java's [JDBC](http://www.oracle.com/technetwork/java/javase/jdbc/index.html), Python's [DB-API](http://www.python.org/dev/peps/pep-0249/), and Microsoft's [ODBC](http://en.wikipedia.org/wiki/ODBC).
It supports the following operations:

* connect/disconnect to the DBMS
* create and execute statements in the DBMS
* extract results/output from statements
* error/exception handling
* information (meta-data) from database objects
* transaction management (optional)

## Installation

Most users who want to access a database do not need to install DBI directly.
It will be installed automatically when you install one of the database backends:

- [RPostgres](https://rpostgres.r-dbi.org) for PostgreSQL,
- [RMariaDB](https://rmariadb.r-dbi.org) for MariaDB or MySQL,
- [RSQLite](https://rsqlite.r-dbi.org) for SQLite,
- [odbc](https://github.com/r-dbi/odbc) for databases that you can access via [ODBC](https://en.wikipedia.org/wiki/Open_Database_Connectivity),
- [bigrquery](https://github.com/r-dbi/bigrquery),
- ... .

You can install the released version of DBI from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("DBI")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("r-dbi/DBI")
```

## Example

The following example illustrates some of the DBI capabilities:

```{r example}
library(DBI)
# Create an ephemeral in-memory RSQLite database
con <- dbConnect(RSQLite::SQLite(), dbname = ":memory:")

dbListTables(con)
dbWriteTable(con, "mtcars", mtcars)
dbListTables(con)

dbListFields(con, "mtcars")
dbReadTable(con, "mtcars")

# You can fetch all results:
res <- dbSendQuery(con, "SELECT * FROM mtcars WHERE cyl = 4")
dbFetch(res)
dbClearResult(res)

# Or a chunk at a time
res <- dbSendQuery(con, "SELECT * FROM mtcars WHERE cyl = 4")
while(!dbHasCompleted(res)){
  chunk <- dbFetch(res, n = 5)
  print(nrow(chunk))
}
dbClearResult(res)

dbDisconnect(con)
```

## Class structure

There are four main DBI classes.
Three which are each extended by individual database backends:

* `DBIObject`: a common base class for all DBI.

* `DBIDriver`: a base class representing overall DBMS properties.
  Typically generator functions instantiate the driver objects like `RSQLite()`,
  `RPostgreSQL()`, `RMySQL()` etc.

* `DBIConnection`: represents a connection to a specific database

* `DBIResult`: the result of a DBMS query or statement.

All classes are _virtual_: they cannot be instantiated directly and instead must be subclassed.

## Further Reading

* [Databases using R](http://db.rstudio.com/) describes the tools and best practices in this ecosystem.

* The [DBI project site](https://www.r-dbi.org/) hosts a blog where recent developments are presented.

* [A history of DBI](https://r-dbi.github.io/DBI/articles/DBI-history.html) by David James, the driving force behind the development of DBI, and many of the packages that implement it.

---

Please note that the _DBI_ project is released with a [Contributor Code of Conduct](https://dbi.r-dbi.org/code_of_conduct).
By contributing to this project, you agree to abide by its terms.
