---
title: "Using DBI with Arrow"
author: "Kirill Müller"
date: "29/09/2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using DBI with Arrow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = Sys.getenv("IN_PKGDOWN") != "true" || (getRversion() < "3.5"))

knit_print.data.frame <- function(x, ...) {
  print(head(x, 6))
  if (nrow(x) > 6) {
    cat("Showing 6 out of", nrow(x), "rows.\n")
  }
  invisible(x)
}

registerS3method("knit_print", "data.frame", "knit_print.data.frame")
```

## Who this tutorial is for

This tutorial is for you if you want to leverage [Apache Arrow](https://arrow.apache.org/) for accessing and manipulating data on databases.
See `vignette("DBI", package = "DBI")` and `vignette("DBI", package = "DBI-advanced")` for tutorials on accessing data using R's data frames instead of Arrow's structures.

## Rationale

Apache Arrow is

> a cross-language development platform for in-memory analytics.

- suitable for large and huge data, also out-of-memory
- data exchange format, good support for data types used in SQL databases
- new extension points to allow backends (currently DuckDB and adbc) to make use of the data exchange format
- faster data retrieval and loading, by avoiding serialization in some cases
- better support for reading and summarizing data from a database that is larger than memory
- better type fidelity with workflows centered around Arrow
- fundamental data structure: `arrow::RecordBatchReader`

## New classes and generics

- Zero chance of interfering with existing DBI backends
- Fully functional fallback implementation for all existing DBI backends
- Requires {arrow} R package

- New generics:
    - `dbReadTableArrow()`
    - `dbCreateTableArrow()`
    - `dbAppendTableArrow()`
    - `dbGetQueryArrow()`
    - `dbSendQueryArrow()`
    - `dbFetchArrow()`
    - `dbFetchArrowChunk()`
    - `dbWriteTableArrow()`

- New classes:
    - `DBIResultArrow`
    - `DBIResultArrowDefault`

## Prepare

```{r}
library(DBI)

con <- dbConnect(RSQLite::SQLite())

data <- data.frame(
  a = 1:3,
  b = 4.5,
  c = "five"
)

dbWriteTable(con, "tbl", data)
```

## Read all rows from a table

```{r}
dbReadTableArrow(con, "tbl")
as.data.frame(dbReadTableArrow(con, "tbl"))
```

## Run queries

```{r}
stream <- dbGetQueryArrow(con, "SELECT COUNT(*) FROM tbl WHERE a < 3")
stream
as.data.frame(stream)
```

## Process data piecemeal

```{r}
stream <- dbGetQueryArrow(con, "SELECT * FROM tbl WHERE a < 3")
stream
stream$get_next()
stream$get_next()
```

## Prepared queries

```{r}
in_arrow <- nanoarrow::as_nanoarrow_array(data.frame(a = 1:4))
stream <- dbGetQueryArrow(con, "SELECT $a AS batch, * FROM tbl WHERE a < $a", param = in_arrow)
as.data.frame(stream)
```

## Writing data

```{r}
stream <- dbGetQueryArrow(con, "SELECT * FROM tbl WHERE a < 3")
dbWriteTableArrow(con, "tbl_new", stream)
dbReadTable(con, "tbl_new")
```

## Appending data

```{r}
stream <- dbGetQueryArrow(con, "SELECT * FROM tbl WHERE a < 3")
dbCreateTableArrow(con, "tbl_split", stream)
dbAppendTableArrow(con, "tbl_split", stream)
stream <- dbGetQueryArrow(con, "SELECT * FROM tbl WHERE a >= 3")
dbAppendTableArrow(con, "tbl_split", stream)
dbReadTable(con, "tbl_split")
```

As usual, do not forget to disconnect from the database when done.

```{r}
dbDisconnect(con)
```

## Conclusion

That concludes the major features of DBI.
For more details on the library functions covered in this tutorial see the DBI specification at `vignette("spec", package = "DBI")`.

- See arrow package for further processing
